---
title: 'Appendix: Evaluating Wine Quality'
author: "Nick Wawee"
date: "12/4/2020"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
rm(list = ls())
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(ggpubr)
library(dplyr)
library(reshape2)
library(LearnEDAfunctions)
library(coin)
library(forcats)
library(jtools)
library(ggstance)
library(broom.mixed)
```

```{r functions, include = F}
plotdists = function(df, path, brtype = 'Scott'){
  plotlist = list()
  for (col in colnames(df)){
    x = df[,col]
    if (class(x)=='factor'){
      dfnew = data.frame(col = factor(), count = integer())
      for (level in levels(x)){
        count = length(which(x==level))
        dfnew = rbind(data.frame(col= level, count = count), dfnew)
      }
      dfnew$col <- factor(dfnew$col, levels = dfnew$col[order(dfnew$count)])
      p = ggplot(dfnew, aes(x=col, y=count))+
        geom_bar(stat= 'identity')+
        plot_opts+
        labs(x = col, y = 'Count')+
        geom_text(aes(label = count), vjust = -0.3) + theme(axis.title = element_text(size = 24))
      plotlist[[col]] = p
    }else if (class(x) == 'numeric' | class(x) == 'integer'){
      dfnew =data.frame(col = class(x))
      histinfo = hist(x = x , breaks=brtype, plot = F)
      p = ggplot(as.data.frame(x), aes(x=x))+
        geom_histogram(bins = length(histinfo$breaks))+plot_opts+
        geom_vline(aes(xintercept = median(x)), 
             linetype = "dashed", size = 1, col = 'red')+
        labs(x = col, y = 'Count')+ theme(axis.title = element_text(size = 24))
        
      plotlist[[col]] = p 
    }
  }
  pfinal =ggarrange(plotlist = plotlist)
  ggsave(paste(path,'dists.pdf',sep=""), pfinal, device = 'pdf',height=1.3*length(plotlist) , width=1.3*length(plotlist)*1.5, units="in", limitsize = FALSE, dpi=900)
  return(pfinal)
}


plot_opts =   theme_bw()+
  theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black") , axis.title.x = element_text(face='bold', size = 14), axis.title.y = element_text(face='bold', size = 14), plot.title = element_text(face='bold', size = 18, hjust=0.5))


#functions
summaries2 = function(datalist){
  letterlist = list()
  outlierlist = list()
  for (name in names(datalist)){
    x = sort(datalist[[name]],decreasing = F) #selecting data
  
    #calculating depths
    depth_m = (length(x) + 1)/2
    depth_f = (depth_m + 1)/2
  
    #calculating letter values
  
    #Mean and Median
    med = median(x)
    average = mean(x)
  
    #Fourth Calculations
    if (depth_m + depth_f %% 1 > 0){
      ind1 = depth_m + depth_f
      ind2 = depth_m - depth_f
      fourth1 = (x[floor(ind1)] + x[ceiling(ind1)])/2
      fourth2 = (x[floor(ind2)] + x[ceiling(ind2)])/2
    }else{
      fourth1 = x[depth_m + depth_f]
      fourth2 = x[depth_m - depth_f]
    }
  
    f_spread = fourth1 - fourth2

    #Max and min
    mx = max(x)
    mn = min(x)
  
  
    #Storing Results for Table
    letterlist[[name]]=data.frame(Letter = c("Maximum", "Upper Fourth", "Median", "Mean", "Lower Fourth", "Minimum", "Outer Upper Fence", "Inner Upper Fence", "Inner Lower Fence", "Outer Lower Fence"), Value = c(mx, fourth1, med, mean(x), fourth2, mn, fourth2 + 2*1.5*f_spread, fourth2 + 1.5*f_spread, fourth1- 1.5*f_spread, fourth1 - 2*1.5*f_spread))
  
    #Finding outliers
    mildind1 = which(x > fourth2 + 1.5*f_spread & x < fourth2 +2*1.5*f_spread)
    mildind2 = which(x < fourth1 - 1.5*f_spread & x > fourth1 - 2*1.5*f_spread)
    mild = x[c(mildind1, mildind2)]
    extreme = x[which(x > fourth2 + 2*1.5*f_spread | x < fourth1 - 2*1.5*f_spread)]
    outlierlist[[name]] = list(mi = mild, ex = extreme)
  }
  return(list(letters = letterlist, outliers = outlierlist))
}#This version of the function does not calculates eighths

minHinkley = function(datalist, powervec){#numeric list and numeric vector
  sampleNames = names(datalist)
  minls=lapply(1:length(sampleNames), function(i){
  
    d = datalist[[sampleNames[i]]]#data
    #calculating depths
    depth_m = (length(d) + 1)/2
    depth_f = (depth_m + 1)/2
    
    #finding minimum for each power
    hvec = unlist(lapply(1:length(powervec), function(j){
      if (powervec[j] != 0){
        d.t = d**powervec[j] # d.t = data transforme
      }else{d.t = log(d)}
      
      x = sort(d.t,decreasing = F) #selecting data
      #Mean and Median
      med = median(x)
      average = mean(x)
  
      #Fourth Calculations
      if (depth_m + depth_f %% 1 > 0){
        ind1 = depth_m + depth_f
        ind2 = depth_m - depth_f
        fourth1 = (x[floor(ind1)] + x[ceiling(ind1)])/2
        fourth2 = (x[floor(ind2)] + x[ceiling(ind2)])/2
        }else{
          fourth1 = x[depth_m + depth_f]
          fourth2 = x[depth_m - depth_f]
        }
      f_spread = fourth1 - fourth2
      
      #Hinkley calculation
      h = (average - med)/f_spread
      
    }))
    abs_hvec = abs(hvec)
    hstar = min(abs_hvec)# minimum distance from 0
    pstar = powervec[which(abs_hvec==hstar)]
    hstar = hvec[which(abs_hvec ==hstar)]
    return(data.frame(min_Hinkley = hstar,Power = pstar))
  })
  names(minls) = names(datalist)
  df = do.call(rbind, minls)
  return(df)
}


```

## Introduction

This document will explore two datasets that comprises physichemical properties and quality ratings of red and white wines. 

```{r loading, include = F}
reddf = read.csv('../data/winequality-red.csv', sep = ';')
whitedf = read.csv('../data/winequality-white.csv', sep = ';')

df = rbind(data.frame(reddf, Type= paste('Red\nN =', as.character(nrow(reddf)))), data.frame(whitedf, Type = paste('White\nN =', as.character(nrow(whitedf)))))
```


## What do the distirubtions look like of both red and white wines?

```{r plot dists, echo = F, message = F}
red_dist = plotdists(reddf, path = '../plots/reddists.png')
white_dist = plotdists(whitedf, path = '../plots/whitedists.png')


df.m = melt(df)
df.m$Type = factor(df.m$Type)

p = ggplot(df.m, aes(x=value, fill = Type))+
  geom_histogram(alpha = 0.5, position = 'identity', bins = 50)+facet_wrap(~variable, scales = 'free')+plot_opts+ labs ( x = 'Value', y = 'Count', title = 'Properties of Red and White Wine')
p
 
ggsave('../plots/dists.png', plot = p, width = 8*4/3, height = 8, units = 'in')
```

\newpage

## What are the letter values of red and white wine?

```{r letter vals, echo = F}
for (col in colnames(whitedf)){
  dlist = list(White = whitedf[,col], Red = reddf[,col])
  s = summaries2(dlist)
  valueslist2=lapply(1:length(s$letters),function(i){ vec= as.data.frame(s$letters[i])[,2]})
  names(valueslist2) = c('White', 'Red')
  combined2 = do.call(cbind,valueslist2)
  row.names(combined2) = s$letters$White$Letter
  print(knitr::kable(combined2, caption = col, digits = 3))
}
```

## What are the characteristics of quality outliers?

### White Wine

There are `r length(s$outliers$White$mi)` mild outliers and `r length(s$outliers$White$ex)` extreme outliers. Below will plot the characteristics of the mild outliers.
\newpage

#### Low Quality

```{r outliers, echo = F}
lowdf = whitedf[whitedf$quality==4,-ncol(whitedf)]
plotdists(lowdf, path = '../plots/white_low_dists.png')
```

\newpage

#### High Quality

```{r outliers2, echo = F}
highdf = whitedf[whitedf$quality>=7,-ncol(whitedf)]
plotdists(highdf, path = '../plots/white_high_dists.png')
```
\newpage

#### Both

```{r bothoutliers, echo = F, message = F}
bothdf = rbind(data.frame(lowdf, Outlier_Type = 'Low'), data.frame(highdf, Outlier_Type = 'High'))
bdf.m = melt(bothdf)

p = ggplot(bdf.m, aes(x=value, fill = Outlier_Type))+
  geom_histogram(alpha = 0.5, position = 'identity', bins = 50)+facet_wrap(~variable, scales = 'free')+plot_opts+ labs ( x = 'Value', y = 'Count', title = 'Outlier Properties of White Wine')+ scale_fill_manual(values = c('#ff0000', '#0000ff'))
p
 
ggsave('../plots/white_outlier_dists.pdf', plot = p, width = 8*4/3, height = 8, units = 'in', device = 'pdf')
```
\newpage

### Red Wine

#### Low Quality

```{r outliers3, echo = F, message = F }
lowdf = reddf[reddf$quality==4,-ncol(reddf)]
plotdists(lowdf, path = '../plots/red_low_dists.png')
```
\newpage

#### High Quality

```{r outliers4, echo = F, message = F }
highdf = reddf[reddf$quality>=7,-ncol(reddf)]
plotdists(highdf, path = '../plots/red_high_dists.png')
```
\newpage

#### Both 

```{r bothoutliers2, echo = F, message = F}
bothdf = rbind(data.frame(lowdf, Outlier_Type = 'Low'), data.frame(highdf, Outlier_Type = 'High'))
bdf.m = melt(bothdf)

p = ggplot(bdf.m, aes(x=value, fill = Outlier_Type))+
  geom_histogram(alpha = 0.5, position = 'identity', bins = 50)+facet_wrap(~variable, scales = 'free')+plot_opts+ labs ( x = 'Value', y = 'Count', title = 'Outlier Properties of Red Wine')+ scale_fill_manual(values = c('#ff0000', '#0000ff'))
p
 
ggsave('../plots/red_outlier_dists.pdf', plot = p, width = 8*4/3, height = 8, units = 'in', device = 'pdf')
```

## What does the symmetry of each variable look like?

### Before Transformation

```{r symmetry, echo = F}
p2 = ggplot(df.m, aes(y = value, x = Type, color = Type))+
  geom_boxplot(outlier.size = 0.1)+ facet_wrap(~variable, scales = 'free_y')+ plot_opts + stat_boxplot(geom= 'errorbar')+ labs(y = 'Value')+ 
  theme(axis.text.x = element_blank())
p2

ggsave('../plots/box.png', plot = p2, width = 8*4/3, height = 8, units = 'in')
```

### Transforming by Optimizing Hinkley Value

```{r minimizing hinkley, echo = F}
redls = list()
for (col in colnames(reddf)){
  redls[[col]] = reddf[,col]
}

redres = minHinkley(redls, seq(from = -2, to = 2, by = 0.01))
knitr::kable(redres, caption = 'Red Wine Results')


whitels = list()
for (col in colnames(whitedf)){
  whitels[[col]] = whitedf[,col]
}

whiteres = minHinkley(whitels, seq(from = -2, to = 2, by = 0.01))
knitr::kable(whiteres, caption = 'White Wine Results')

combinedres = data.frame()
for (r in rownames(whiteres)){
  combinedres[r,'Mean_Power'] = mean(c(whiteres[r,2], redres[r,2]))
}

knitr::kable(combinedres, caption = 'Combined Results', digits = 4)

for (col in colnames(whitedf)){
  meanp = combinedres$Mean_Power[rownames(combinedres)==col]
  if (meanp == 0){
    whitedf[,col] == log(whitedf[,col])
    reddf[,col] == log(reddf[,col])
  }else{
  whitedf[,col] = whitedf[,col]**meanp
  reddf[,col] = reddf[,col]**meanp
  }
}

```

### Plotting Transformed Variables

```{r symmetry2, echo = F, message = F}

df = rbind(data.frame(reddf, Type= paste('Red\nN =', as.character(nrow(reddf)))), data.frame(whitedf, Type = paste('White\nN =', as.character(nrow(whitedf)))))

df.m = melt(df)
df.m$Type = factor(df.m$Type)

p2 = ggplot(df.m, aes(y = value, x = Type, color = Type))+
  geom_boxplot(outlier.size = 0.1)+ facet_wrap(~variable, scales = 'free_y')+ plot_opts + stat_boxplot(geom= 'errorbar')+ labs(y = 'Value')+ 
  theme(axis.text.x = element_blank())+labs(y = 'Value^mean(p)')
p2

ggsave('../plots/box_transform.png', plot = p2, width = 8*4/3, height = 8, units = 'in')
```

```{r untransform, echo = F}
reddf = read.csv('../data/winequality-red.csv', sep = ';')
whitedf = read.csv('../data/winequality-white.csv', sep = ';')
df = rbind(data.frame(reddf, Type= 'Red'), data.frame(whitedf, Type = 'White'))
```

## Which variables are statistically different between the untransformed data?

Below will employ a two sample t test to test if the populations are equal for each variable. We assume unequal variances despite the power transformation as the IQR spread is not equal in all cases. We will also run a Mood's Median test to test for differences in median rather than mean.

```{r ttest_mood, echo = F}
tdf = data.frame()
mdf = data.frame()
for (col in colnames(whitedf)){
  #T test
  res = t.test(x = whitedf[,col], y = reddf[,col])
  tdf[col, 't'] = res$statistic
  tdf[col, 'p'] = res$p.value
  
  #Mood's median test
  df.s = df.m[df.m$variable==col,]
  medres = median_test(value~Type, data = df.s)
  
  mdf[col, 'Z'] = medres@statistic@teststatistic
  mdf[col, 'p'] = 2*pnorm(-abs(medres@statistic@teststatistic))
}
tdf = tdf[order(tdf$t),]
mdf = mdf[order(mdf$Z),]
knitr::kable(tdf, caption = 't-test Results', digits = 3)
knitr::kable(mdf, caption = "Mood's Median Results", digits = 3)
```

## Which factors impact the quality score the most?

### Examining Correlations with Quality

```{r correlations, echo = F, message = F}
wcordf = data.frame()
rcordf = data.frame()
meancor = data.frame()
for (col in colnames(whitedf[,-ncol(whitedf)])){
  wcordf[col, 'r'] = cor(whitedf[,col], whitedf[,'quality'])
  rcordf[col, 'r'] = cor(reddf[,col], reddf[,'quality'])
  meancor[col, 'r'] = mean(c(cor(reddf[,col], reddf[,'quality']), cor(whitedf[,col], whitedf[,'quality'])))
}
wcordf$Type = 'White'
rcordf$Type = 'Red'
rcordf$variable = rownames(rcordf)
wcordf$variable = rownames(wcordf)
meancor = as.matrix(meancor)
meancor = meancor[order(meancor),]
meancor = data.frame(r = meancor)
meancor$o = seq(from = 1, to = nrow(meancor), by = 1)

rcordf$o[match(rownames(meancor), rcordf$variable)] = meancor$o
wcordf$o[match(rownames(meancor), wcordf$variable)] = meancor$o

cordf = rbind(wcordf, rcordf)
#cordf = cordf[order(cordf$o),]
cordf$Type = as.factor(cordf$Type)
cordf$variable = as.factor(cordf$variable)
p3 = cordf %>%
  mutate(variable = fct_reorder(variable, o)) %>%
  ggplot(aes(x=variable, y = r))+
           geom_dotplot(binaxis = "y", stackdir = "center", aes( fill= Type))+plot_opts+theme(axis.text.x = element_text(angle = 45, vjust = .6))+
    labs(x = 'Variable')

p3

ggsave('../plots/correlations.png', plot = p3, height = 4, width = 5, units = 'in', dpi = 600)
```

### MLR


#### Modeling

```{r ols, echo = F}
w_mlr = lm(quality ~ 0+ fixed.acidity +volatile.acidity+ citric.acid+residual.sugar+ chlorides+ free.sulfur.dioxide + total.sulfur.dioxide + density+ pH+ sulphates + alcohol, data = whitedf)
summary(w_mlr)

coefdf = as.data.frame(summary(w_mlr)$coefficients)
coefdf = coefdf[order(coefdf$Estimate),]
knitr::kable(coefdf, digits = 3, caption = 'White Wine Regression Coefficients')

r_mlr = lm(quality ~ 0+ fixed.acidity +volatile.acidity+ citric.acid+residual.sugar+ chlorides+ free.sulfur.dioxide + total.sulfur.dioxide + density+ pH+ sulphates + alcohol, data = reddf)
summary(r_mlr)

coefdf = as.data.frame(summary(r_mlr)$coefficients)
coefdf = coefdf[order(coefdf$Estimate),]
knitr::kable(coefdf, digits = 3, caption = 'Red Wine Regression Coefficients')
```


#### Plotting Coefficients

```{r plot coefs, echo = F}
cs = c('#00BFC4', '#F8766D')
pc = plot_coefs(w_mlr, r_mlr, model.names = c('White','Red'), colors = cs)+theme(axis.text.y = element_text(face = 'bold'))

pc
ggsave('../plots/coefplot.png', plot = pc, height = 7, width = 7, dpi = 600, units = 'in')
```

